Robust inference of neutron-star parameters from thermonuclear burst

### 2019-08-28 doc/ms.tex
Updated paper

\newcommand{\eps}{{\rm erg\,s^{-1}}}
\newcommand{\epcs}{{\rm erg\,cm^{-2}\,s^{-1}}}
\newcommand{\epc}{{\rm erg\,cm^{-2}}}

@@

Thermonuclear bursts arise from unstable ignition of accumulated fuel on the surface of neutron stars in low-mass X-ray binaries. Comparison of observed bursts with numerical models is a key factor in determining the properties of the host neutron star.
%
Here we describe some commonly-used approaches for determining system parameters, including composition of the burst fuel, and introduce a new suite of software tools, {\sc concord}, intended to make comprehensive model-observation comparisons practical.

@@

@@

The fundamental observable that we measure is the time-history of the burst flux, the lightcurve $(t_i,F_{i})$. Here the $F_{i}$ may be in instrumental units of count~s$^{-1}$, or may be estimates of the bolometric flux at time $t_i$, derived from model fits of the spectrum at time $t_i$.
%
Provided the $F_i$ are in units of flux, or the flux can be estimated from the values,  we may measure the peak flux, $F_{\rm pk}$, and integrate to estimate the burst fluence, $E_b$.

The peak flux $F_{\rm pk}$ provides constraints on the source distance, since the maximum burst luminosity is limited to (roughly) the Eddington luminosity \cite[e.g.][]{lew93}
% copied from bcatalog
\begin{eqnarray}
  L_{\rm Edd,\infty} & = & \frac{8\pi G m_p M_{\rm NS} c
  [1+(\alpha_{\rm T}T_{\rm e})^{0.86}]} {\sigma_T(1+X)[1+z(R)]} % \xi_b
       \nonumber \\
  & = & 2.7\times10^{38} \left(\frac{M_{\rm NS}}{1.4M_\odot}\right)
 \frac{1+(\alpha_{\rm T}T_{\rm e})^{0.86}}{(1+X)}
\nonumber \\ & & \times\  % line break
    \left[\frac{1+z(R)}{1.31}\right]^{-1}\ % \xi_b\
              \eps
  \label{ledd}
\end{eqnarray}
Here
$T_{\rm e}$ is the
effective temperature of the atmosphere, $\alpha_{\rm T}$ parametrizes the temperature dependence of the electron scattering
opacity \cite[$\simeq 2.2\times10^{-9}$~K$^{-1}$;][]{lew93},
$m_p$ is the proton mass, $\sigma_T$ the Thompson 
cross-section, and $X$ the
mass fraction of hydrogen in the atmosphere.
% ($\approx0.7$ for cosmic abundances).
The final factor in
square brackets represents the gravitational redshift 
at the photosphere $1+z(R)=(1-2GM_{\rm NS}/R
c^2)^{-1/2}$, which may be lower than the value at the NS surface, while the photosphere is expanded during the radius expansion episode (i.e. $R\ge R_{\rm NS}$).
%
Mention also the empirical value here...

A significant fraction of observed bursts exhibit characteristic variations in their spectral hardness (or the blackbody temperature and normalisation) around their maximum that indicate the presence of photospheric radius-expansion (PRE). By equating the maximum flux of these events with the Eddington luminosity, the distance to the source can be estimated as
\begin{eqnarray}
 d & = & \left(\frac{L_{\rm Edd,\infty}}{4\pi \xi_b F_{\rm pk, RE}}\right)^{1/2} \nonumber \\
   & = & 8.6
	\left( \frac{\xi_b F_{\rm pk, RE}}{3\times10^{-8}\ \epcs} \right)^{-1/2}
       	\left(\frac{M_{\rm NS}}{1.4M_\odot}\right)^{1/2}
\nonumber \\ & & \times\  % line break
	\left[\frac{1+z(R)}{1.31}\right]^{-1/2}
	(1+X)^{-1/2}\ {\rm kpc}
 \label{disteq}
\end{eqnarray}
% 
Here the $\xi_b$ factor accounts for the possible anisotropy of the burst emission, and is defined in the same sense as \citep{fuji88}, such that $L_b=4\pi d^2\xi_b F_{\rm pk}$. Since the distribution of the burst flux over the neutron star is likely different from that of the persistent emission, we define a different anisotropy factor $\xi_p$ relating the observed persistent flux $F_p$ to the total luminosity. Modelling \cite[]{he16}...

@@

The fluence (and the shape of the burst profile) is determined by the amount of accumulated fuel, and its composition. The nuclear energy generation rate $Q_{\rm nuc}$ depends on the composition; studies with the 1-D numerical code {\sc kepler} suggest that 
\begin{eqnarray}
Q_{\rm nuc} &= & 1.31+6.95\bar{X}-1.92\bar{X}^2 \nonumber \\
& \approx & 1.35 + 6.05 \bar{X}
\end{eqnarray}
\cite[]{goodwin19a} where $\bar{X}$ is the average H-fraction of the fuel layer at ignition.

Perhaps put in here some estimates of the fuel composition as a function of $X_0$ and recurrence time; could perhaps develop a ``prior'' for $\bar{X}$

Provided an estimate of $\bar{X}$ is available, we can estimate the burst column that has ignited as
\begin{eqnarray}
y & = & \frac{L_b d^2 (1+z)}{R_{\rm NS}^2 Q_{\rm nuc}}\ 
                                                            \nonumber \\
  & = & 3.0\times10^8 \left(\frac{\xi_b E_{\rm b}}{10^{-6}\ \epc}\right)
                      \left(\frac{d}{10\ {\rm kpc}}\right)^2\
             \left(\frac{Q_{\rm nuc}}{4.4\ {\rm MeV/nucleon}}\right)^{-1}
\nonumber \\ & & \times\  % line break
                      \left(\frac{1+z}{1.31}\right)
                      \left(\frac{R_{\rm NS}}{10\ {\rm km}}\right)^{-2}
                      {\rm g\,cm^{-2}}
\label{column}
\end{eqnarray}
{\it need to update prefactor, denominator in $Q_{\rm nuc}$ term here }

@@

An additional correction that must be applied is the effect of the anisotropic emission of the burst, based on the system inclination $i$. This effect has been simulated for a range of disk geometries \cite[e.g.][]{he16}. Because the emission region on the surface of the star is thought to be different for the burst and persistent emission, we define separate anisotropy parameters $\xi_b$ and $\xi_p$, with the sense that
\begin{equation}
L_{b,p} = 4\pi d^2\xi_{b,p}F_{b,p}
\end{equation}
where $L_b$, $L_p$ are the burst and persistent luminosity, $F_b$, $F_p$ are the corresponding (observed) fluxes, and $d$ is the source distance. With this sense it can be seen that values of $\xi_{b,p}<1$ correspond to emission preferentially beamed towards us, so that --- without the anisotropy corrections --- the isotropic luminosity inferred from the observed flux would overestimate the actual luminosity.

@@

Here we describe three different approaches that have been used to deduce system parameters from observations of thermonuclear bursts.

\subsection{Inferring fuel composition} 
\label{subsec:fuelcomp}

Given a set of burst observations as described in \S\ref{subsec:data}, a common approach \cite[e.g][]{falanga11} is to estimate the fuel composition at ignition, and hence the accreted fuel composition, based on simple analytic estimates of the burst energy production $Q_{\rm nuc}$.

One issue with this approach is that the effects of anisotropy are frequently neglected.
%
A second issue which has emerged recently is that the commonly used estimates for the nuclear energy generation rate $Q_{\rm nuc}$ are not supported by recent {\sc kepler} models. 
%
We describe the general approach below including the effects of these two issues.

>>>>

\newcommand{\eps}{{\rm erg\,s^{-1}}}
\newcommand{\epcs}{{\rm erg\,cm^{-2}\,s^{-1}}}
\newcommand{\epc}{{\rm erg\,cm^{-2}}}
\newcommand{\Xb}{\ensuremath{\overline{X}}}

@@

Thermonuclear (type-I) bursts arise from unstable ignition of accumulated fuel on the surface of neutron stars in low-mass X-ray binaries. Measurements of burst properties enable observers to infer the properties of the host neutron stars, but a number of confounding astrophysical effects contribute to systematic uncertainties. 
%
In particular, comparison of observed burst properties with the predictions of numerical models is a 
% key factor in determining the properties of the host neutron star.
powerful method of constraining host properties.
%
Here we describe some commonly-used approaches for determining system parameters, including composition of the burst fuel, and introduce a new suite of software tools, {\sc concord}, intended to make comprehensive model-observation comparisons practical.

@@

Thermonuclear (type-I) bursts are observed from neutron stars accreting from low-mass ($\lesssim 1\ M_\odot$) binary companions \cite[e.g.][]{gal17b}. ...

@@

The fundamental observable that we measure is the time-history of the burst flux, the lightcurve $(t_i,F_{i})$. Here the $F_{i}$ may be in instrumental units of count~s$^{-1}$, or may be estimates of the bolometric flux at time $t_i$, derived from model fits of the spectrum at time $t_i$.
%
Provided the $F_i$ are in units of flux, or the flux can be estimated from the intensity,  we may measure the peak flux, $F_{\rm pk}$, and integrate to estimate the burst fluence, $E_b$.

The peak flux $F_{\rm pk}$ provides constraints on the source distance, since the maximum burst luminosity is limited to (roughly) the Eddington luminosity \cite[e.g.][]{lew93}
% copied from bcatalog, and updated prefac
% (8.*pi*const.G*const.m_p*1.4*const.M_sun*const.c/(const.sigma_T*1.259)).to('erg s-1')                                                         
% Out[16]: <Quantity 2.79569695e+38 erg / s>
\begin{eqnarray}
  L_{\rm Edd,\infty} & = & \frac{8\pi G m_p M_{\rm NS} c
  [1+(\alpha_{\rm T}T_{\rm e})^{0.86}]} {\sigma_T(1+X)[1+z(R)]} % \xi_b
       \nonumber \\
%  & = & 2.7\times10^{38} \left(\frac{M_{\rm NS}}{1.4M_\odot}\right)
  & = & 2.80\times10^{38} \left(\frac{M_{\rm NS}}{1.4M_\odot}\right)
 \frac{1+(\alpha_{\rm T}T_{\rm e})^{0.86}}{(1+X)}
\nonumber \\ & & \times\  % line break
%    \left[\frac{1+z(R)}{1.31}\right]^{-1}\ % \xi_b\
    \left[\frac{1+z(R)}{1.259}\right]^{-1}\ % \xi_b\
              \eps
  \label{ledd}
\end{eqnarray}
Here
$T_{\rm e}$ is the
effective temperature of the atmosphere, $\alpha_{\rm T}$ parametrizes the temperature dependence of the electron scattering
opacity \cite[$\simeq 2.2\times10^{-9}$~K$^{-1}$;][]{lew93},
$m_p$ is the proton mass, $\sigma_T$ the Thompson 
cross-section, and $X$ the
mass fraction of hydrogen in the atmosphere.
% ($\approx0.7$ for cosmic abundances).
The final factor in
square brackets represents the gravitational redshift 
at the photosphere $1+z(R)=(1-2GM_{\rm NS}/R
c^2)^{-1/2}$, which may be lower than the value at the NS surface, while the photosphere is expanded during the radius expansion episode (i.e. $R\ge R_{\rm NS}$).

% {\it Mention also the empirical value here... } actually later

A significant fraction of observed bursts exhibit characteristic variations in their spectral hardness (or the blackbody temperature and normalisation) around their maximum that indicate the presence of photospheric radius-expansion (PRE). By equating the maximum flux of these events with the Eddington luminosity, the distance to the source can be estimated as
% p is the prefactor from the L_Edd expression above
% np.sqrt(p/(4.*pi*3e-8*u.erg/u.cm**2/u.s)).to('kpc')                     
% Out[26]: <Quantity 8.82528911 kpc>
\begin{eqnarray}
 d & = & \left(\frac{L_{\rm Edd,\infty}}{4\pi \xi_b F_{\rm pk, RE}}\right)^{1/2} \nonumber \\
%   & = & 8.6
   & = & 8.83
	\left( \frac{\xi_b F_{\rm pk, RE}}{3\times10^{-8}\ \epcs} \right)^{-1/2}
       	\left(\frac{M_{\rm NS}}{1.4M_\odot}\right)^{1/2}
\nonumber \\ & & \times\  % line break
%	\left[\frac{1+z(R)}{1.31}\right]^{-1/2}
	\left[\frac{1+z(R)}{1.259}\right]^{-1/2}
	(1+X)^{-1/2}\ {\rm kpc}
 \label{disteq}
\end{eqnarray}
% 
Here the $\xi_b$ factor accounts for the possible anisotropy of the burst emission, 
which is discussed in the next section.

@@

\subsection{Emission anisotropy}
\label{subsec:anisotropy}

The anisotropy factor $\xi_b$ introduced into the expression for the distance (equation  \ref{disteq}) is defined in the same sense as \citep{fuji88}, such that the total luminosity of the burst $L_b=4\pi d^2\xi_b F_{\rm pk}$. The range of values of $\xi_b$ can thus be understood that if $\xi_b<1$ the burst flux is enhanced (i.e. preferentially beamed) toward our line of sight (so that $L_b$ would be {\it overestimated} were it not included in the calculation), while if $\xi_b>1$ the burst flux is suppressed.
%
Since the distribution of the burst flux over the neutron star is likely different from that of the persistent emission, we define a different anisotropy factor $\xi_p$ relating the observed persistent flux $F_p$ to the total luminosity. Modelling \cite[]{he16}...

{\it describe how we treat this }

\subsection{Burst energetics}
\label{subsec:energetics}

The fluence (and the shape of the burst profile) is determined by the amount of accumulated fuel, and its composition. The nuclear energy generation rate $Q_{\rm nuc}$ depends on the composition; studies with the 1-D numerical code {\sc kepler} suggest that 
\begin{eqnarray}
Q_{\rm nuc} &= & 1.31+6.95\Xb-1.92\Xb^2 \nonumber \\
& \approx & 1.35 + 6.05 \Xb
\end{eqnarray}
\cite[]{goodwin19a} where $\Xb$ is the average H-fraction of the fuel layer at ignition. For fuel with solar composition (i.e. $X=0.7$), $Q_{\rm nuc}=5.22$~MeV/nucleon.

{\it Perhaps put in here some estimates of the fuel composition as a function of $X_0$ and recurrence time; could perhaps develop a ``prior'' for $\Xb$ }

Provided an estimate of $\Xb$ is available, we can estimate the burst column that has ignited as
% (1e-6*u.erg/u.cm**2*(10.*u.kpc).to('cm')**2/(5.22*u.MeV/const.m_p)*1.259
%     ...: /(11.2*u.km)**2).to('g cm-2')
% Out[36]: <Quantity 1.9112067e+08 g / cm2>
\begin{eqnarray}
y & = & \frac{L_b d^2 (1+z)}{R_{\rm NS}^2 Q_{\rm nuc}}\ 
                                                            \nonumber \\
%  & = & 3.0\times10^8 \left(\frac{\xi_b E_{\rm b}}{10^{-6}\ \epc}\right)
  & = & 1.91\times10^8 \left(\frac{\xi_b E_{\rm b}}{10^{-6}\ \epc}\right)
                      \left(\frac{d}{10\ {\rm kpc}}\right)^2\
%             \left(\frac{Q_{\rm nuc}}{4.4\ {\rm MeV/nucleon}}\right)^{-1}
             \left(\frac{Q_{\rm nuc}}{5.22\ {\rm MeV/nucleon}}\right)^{-1}
\nonumber \\ & & \times\  % line break
%                      \left(\frac{1+z}{1.31}\right)
%                      \left(\frac{R_{\rm NS}}{10\ {\rm km}}\right)^{-2}
                      \left(\frac{1+z}{1.259}\right)
                      \left(\frac{R_{\rm NS}}{11.2\ {\rm km}}\right)^{-2}
                      {\rm g\,cm^{-2}}
\label{column}
\end{eqnarray}
% {\it need to update prefactor, denominator in $Q_{\rm nuc}$ term here } - done

@@

{\it This likely duplicates earlier discussion} \\
%
An additional correction that must be applied is the effect of the anisotropic emission of the burst, based on the system inclination $i$. This effect has been simulated for a range of disk geometries \cite[e.g.][]{he16}. Because the emission region on the surface of the star is thought to be different for the burst and persistent emission, we define separate anisotropy parameters $\xi_b$ and $\xi_p$, with the sense that
\begin{equation}
L_{b,p} = 4\pi d^2\xi_{b,p}F_{b,p}
\end{equation}
where $L_b$, $L_p$ are the burst and persistent luminosity, $F_b$, $F_p$ are the corresponding (observed) fluxes, and $d$ is the source distance. With this sense it can be seen that values of $\xi_{b,p}<1$ correspond to emission preferentially beamed towards us, so that --- without the anisotropy corrections --- the isotropic luminosity inferred from the observed flux would overestimate the actual luminosity.

@@

Here we describe different approaches that have been used to deduce system parameters from observations of thermonuclear bursts.
%
We focus on adopting a suitable approach given the availability of data, which will differ for different sources

\subsection{Zero bursts -- constraining the distance}
\label{subsec:zerobursts}

While it may seem strange to base an analysis on the {\it non-}detection of bursts, practically even for the best-studied sources the X-ray observation duty cycle is of order a few percent. There is a high probability that bursts will be missed, particularly if the accretion rate is low (and hence the bursts are infrequent). In extreme cases this can mean that no bursts whatsoever are observed, which may instead be explained by the compact object being a black hole rather than a neutron star.

In cases where we can be confident that the compact object is a neutron star (e.g. where persistent pulsations are detected) but no bursts are observed, we can constrain the distance by adopting a composition for the fuel and comparing the predictions of burst models to the good-time intervals of our X-ray data.

For example, based on {\it RXTE}/PCA observations of the accretion-powered millisecond pulsar IGR~J00291+5934, \cite{gal06b} derived joint constraints over the distance $d$ and fuel H-fraction $X_0$, adopting  the predictions of a simple numerical model ({\it ref to {\sc settle}}). These constraints suggest the distance is $\gtrsim5$~kpc (at 3-sigma significance), provided $X_0\approx0.7$, a reasonable choice given the expectation of a H-rich donor in this $2.46$~hr binary orbital period system \cite[]{gal05a}.

In a subsequent outburst in 2015, a single burst was detected by {\it Swift}/XRT \cite[]{kuin15}, offering the opportunity to verify the previously-determined distance limit. The estimated peak 
% 0.3--10~keV flux in the range (3.0--$5.6)\times10^{-11}\ \epcs$. 
0.1--35~keV flux was $(18\pm4)\times10^{-8}\ \epcs$ \cite[]{defalco17}, with the burst exhibiting spectral variations indicative of PRE. By comparing the peak flux to the empirical Eddington luminosity, the corresponding distance 
\cite[including the inclination range of 22--32$^\circ$ suggested by][]{torres08},
% possible inclinations as described in sec \S\ref{subsec:data}) is $4.4_{-0.7}^{+0.8}$~kpc, which is mostly inconsistent with the earlier limit. Clearly, such limits must be viewed with some caution.
is $5.0_{-0.5}^{+0.7}$~kpc, which is fully consistent with the previously established limit. The effect of the inclination constraints is illustrated in Fig. \ref{fig:dist_igr00291}).

\begin{figure}[ht!]
\plotone{fig1.pdf}
\caption{Estimated distance distributions for the accretion-powered X-ray pulsar IGR~J00291+5934, based on the peak flux of the sole thermonuclear burst observed by \cite{defalco17}.
%
The distribution for an isotropic distribution of system inclinations (up to a maximum value of $75^\circ$, motivated by the lack of dips in the X-ray intensity) is shown as the blue shaded histogram. Imposing the inclination constraint of 22--32$^\circ$ suggested by \cite{torres08}, for which the predicted burst anisotropy factor is $\xi_b=0.704$ on average, gives instead the orange-shaded histogram. The resulting $1\sigma$ confidence interval is $5.0_{-0.5}^{+0.7}$~kpc.
\label{fig:dist_igr00291}}
\end{figure}

\subsection{One burst -- estimating the burst rate} 
\label{subsec:burstrate}

{\it copy the analysis for the IGR source here }

\subsection{Two bursts -- inferring fuel composition} 
\label{subsec:fuelcomp}

Given a minimally complete set of burst observations as described in \S\ref{subsec:data}, a common approach \cite[e.g][]{falanga11} is to estimate the fuel composition at ignition, and hence the accreted fuel composition, based on simple analytic estimates of the burst energy production $Q_{\rm nuc}$.

One issue with this approach (as for analyses based on one burst, as described in \S\ref{subsec:burstrate}) is that the effects of anisotropy are frequently neglected.
%
A second issue which has emerged recently is that the commonly used estimates for the nuclear energy generation rate $Q_{\rm nuc}$ are not supported by recent {\sc kepler} models. 
%
We describe the general approach below including the effects of these two issues.
